Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-20 _tex/index.tex
Built site for gh-pages

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(T\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((T=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed by the
associational random quantity \(E[Y| T=1] - E[Y| T=2]\), where
\(E[\cdot]\) denotes the expected value. An example of this approach is
seen in \citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that maps data into an estimate \citep{Neal_2020}. \emph{Estimates} are
numerical values that approximate the estimand and are derived through
\emph{estimation}, which refers to the process of integrating data with
an estimator \citep{Everitt_et_al_2010}. Although various methods can
approximate an estimand, researchers prioritize estimators with
desirable properties that ensure the accuracy of estimates. For
instance, the Z-test is an estimator known for its effectiveness in
comparing two proportions, yielding accurate estimates when its
underlying assumptions are met \citep{Kanji_2006}. The Z-test is
expressed as a signal-to-noise ratio:
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is the
difference between the group sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y| T=1]\) and \(E[Y| T=2]\), respectively. The noise \(\hat{s}_{p}\)
is the unpooled sample variability observed between the two groups.

@@

Experiments are widely recognized as the gold standard in evidence-based
science \citep{Hariton_et_al_2018, Hansson_2014}. However, researchers
often face constraints that limit their ability to conduct experimental
studies. These constraints include ethical concerns, such as the
assignment of individuals to potentially harmful interventions, and
practical limitations, such as the infeasibility of, for example,
assigning individuals to genetic modifications or physical impairments
\citep{Neal_2020}. In these situations, causal inference provides a
valuable alternative for generating causal estimates, particularly when
the goal is to understand the mechanisms underlying specific data.
Moreover, the framework offers significant theoretical insights that
enhance the design of observational and experimental studies
\citep{McElreath_2020}.

Unlike classical statistical modeling, which focuses primarily on
summarizing data and inferring associations, \emph{causal inference} is
a framework designed to identify causes and estimate their effects using
data \citep{Shaughnessy_et_al_2010, Neal_2020}. This framework employs
rigorous mathematical techniques to address the fundamental problem of
causality \citep{Pearl_2009}, which is essential for understanding and
defining causal effects. The \emph{fundamental problem of causality}
centers on answering the question, ``What would have happened `in the
world' under different circumstances?'' a concept known as
counterfactuals. \emph{Counterfactuals} represent hypothetical scenarios
that are \emph{contrary to fact}, where alternative outcomes resulting
from a specific cause are neither observed nor observable
\citep{Neal_2020, Counterfactual_2024}.

Although a comprehensive discussion of causes and counterfactuals
exceeds the scope of this document, a brief overview of how the
framework addresses the fundamental problem of causality is possible
using the previous example. The framework starts by defining
\emph{individual causal effects} (ICE) as the difference between
students' observed and unobserved potential outcomes:
\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice once a
student is assigned to \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) is no longer observed nor observable, and is thus termed a
\emph{counterfactual}. To address the challenge posed by
counterfactuals, the framework extends the ICE to \emph{average causal
effects} (ACE). The ACE is defined as
\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\), and
represents the difference between the average of observed potential
outcomes and counterfactuals in the sample. Finally, akin to
experimental studies, the ACE is identified from associational estimates
by ensuring the absence of confounding. This is accomplished by
statistically conditioning on a \emph{sufficient adjustment set} of
variables (\(X\)). Consequently, the ACE is expressed as
\(\tau = E_{X}[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]]\), where
\(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

@@

The structural approach, however, offers a notable advantage over the
potential outcomes approach by allowing the graphical and formal
representation of the ACE through directed acyclic graphs
\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
DAGs function as heuristics, effectively conveying the presumed causal
structure of a system, referred to as a \emph{scientific model}. They do
not represent detailed statistical models but allow researchers to
deduce which statistical models can provide valid causal inferences,
assuming the causal structure depicted in the DAGs are accurate
\citep{McElreath_2020}. The Identification-Estimation flowchart in
Figure~\ref{fig-IEflow} visually represents the process of transitioning
from estimands to estimates, as well as the application of the
scientific model and data to identify and estimate causal effects.

@@

DAGs offer two key advantages for modeling causal structures. Firstly,
they represent causal relations in a nonparametric and fully interactive
manner. This feature allows for feasible causal analysis strategies
without needing the specification of the type of data or the nature of
the functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature enables the
decomposition of complex structures into basic building blocks,
facilitating the analysis of these structures by focusing on the causal
assumptions associated with each building block individually
\citep{McElreath_2020}. These building blocks can be represented in
three ways: the magnified representation, the standard representation,
and the structural causal model form \citep[SCM,][]{Morgan_et_al_2014}.

@@

The motivating example in Section~\ref{sec-framework-structural} can
further illustrate how to use the five fundamental building blocks to
construct a system's causal structure. In this scenario, the
investigator aims to determine whether, \emph{on average}, there is a
difference in conceptual understanding of a topic \((Y=\{0,1\})\)
between two groups of students \((T=\{1,2\})\), described by the
estimand \(E[Y| T=1] - E[Y| T=2]\). However, unlike the previous case,
an experiment cannot be conducted, and the problem suggests that the
country to which a student belongs (\(X\)) may influence both \(T\) and
\(Y\). Such scenarios are plausible, especially when the teaching
methods depend on software or access to technology, which may be limited
in certain countries {(maybe an example with more impact?)}.
Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate the
fork \(X \rightarrow \{T,Y\}\), and two colliders with
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

>>>>

Empirical research addresses real-world challenges by relying on
evidence gathered through observation and experimentation. In this
context, researchers typically frame their research questions as
\emph{estimands} or \emph{targets of inference}. These estimands
represent the specific quantities the study aims to determine
\citep{Everitt_et_al_2010}. For instance, a study might examine the
question, ``To what extent do different teaching methods (\(T\))
influence students' conceptual understanding of a topic (\(Y\))?'' To
investigate this, the study could randomly assign students to two
groups, each using a different teaching method \((T=\{1,2\})\).
Students' conceptual understanding of the topic could be evaluated
through pairwise comparisons, resulting in a dichotomous outcome
\((Y=\{0,1\})\), indicating which student among those compared has a
higher level of understanding. The research question could be then
framed as the estimand, ``\emph{On average}, is there a difference in
conceptual understanding of the topic between the two groups of
students?'' This estimand could be mathematically expressed by the
random quantity \(E[Y| T=1] - E[Y| T=2]\), where \(E[\cdot]\) denotes
the expected value. An example of this approach is seen in
\citet{Jones_et_al_2019}.

Researchers then proceed to identify the estimands.
\emph{Identification} refers to the process of accurately computing an
estimand using an estimator. An \emph{estimator} is a method or function
that transforms data into an estimate \citep{Neal_2020}.
\emph{Estimates} are numerical values that approximate the estimand and
are derived through \emph{estimation}, which refers to the process of
integrating data with an estimator \citep{Everitt_et_al_2010}. Although
various methods can approximate an estimand, researchers prioritize
estimators with desirable properties that ensure the accuracy of
estimates. For instance, the Z-test is an estimator known for its
effectiveness in comparing groups' proportions, yielding accurate
estimates when the underlying assumptions of the statistic are met
\citep{Kanji_2006}. The Z-test is expressed as a signal-to-noise ratio:
\(Z = (\hat{p}_{1} - \hat{p}_{2})/ \hat{s}_{p}\). The signal is the
difference between the group sample proportions,
\(\hat{p}_{1} = \sum_{i=1}^{n_{1}}{Y_{i}/n_{1}}\) and
\(\hat{p}_{2} = \sum_{i=1}^{n_{2}}{Y_{i}/n_{2}}\), analogous to
\(E[Y| T=1]\) and \(E[Y| T=2]\), respectively. The noise \(\hat{s}_{p}\)
is the unpooled sample variability observed between the two groups.

@@

Experiments are widely recognized as the gold standard in evidence-based
science \citep{Hariton_et_al_2018, Hansson_2014}. However, researchers
often face constraints that limit their ability to conduct experimental
studies. These constraints include ethical concerns, such as the
assignment of individuals to potentially harmful interventions, and
practical limitations, such as the infeasibility of, for example,
assigning individuals to genetic modifications or physical impairments
\citep{Neal_2020}. In these situations, causal inference provides a
valuable alternative for generating causal estimates, particularly when
the goal is to understand the mechanisms underlying specific data.
Moreover, the framework provides significant theoretical insights that
enhance the design of observational and experimental studies
\citep{McElreath_2020}.

Unlike classical statistical modeling, which focuses primarily on
summarizing data and inferring associations, \emph{causal inference} is
a framework designed to identify causes and estimate their effects using
data \citep{Shaughnessy_et_al_2010, Neal_2020}. The framework employs
rigorous mathematical techniques to address the \emph{fundamental
problem of causality} \citep{Pearl_2009}. This problem revolves around
the question, ``What would have happened `in the world' under different
circumstances?'' a concept known as counterfactuals, essential for
understanding and defining causal effects. \emph{Counterfactuals}
represent hypothetical scenarios that are \emph{contrary to fact}, where
alternative outcomes resulting from a specific cause are neither
observed nor observable \citep{Neal_2020, Counterfactual_2024}.

Although a comprehensive discussion of causes and counterfactuals
exceeds the scope of this document, a brief overview of how the
framework addresses the fundamental problem of causality can be
provided. The framework begins by defining \emph{individual causal
effects} (ICE) as the difference between students' observed and
unobserved potential outcomes:
\(\tau_{i} = (Y_{i}| T_{i}=1) - (Y_{i}| T_{i}=2)\). Notice when a
student is assigned to \(T_{i}=1\), the potential outcome under
\(T_{i}=2\) is no longer observed nor observable, and thus termed a
\emph{counterfactual}. To overcome the challenge posed by
counterfactuals, the framework extends the ICE to \emph{average causal
effects} (ACE). The ACE is defined as
\(\tau = E[\tau_{i}] = E[Y_{i}| T_{i}=1] - E[Y_{i}| T_{i}=2]\),
representing the difference between the average of observed potential
outcomes and counterfactuals across the sample. Finally, similar to
experimental studies, the counterfactuals are identified from
associational estimates by ensuring the absence of confounding. This is
accomplished by statistically conditioning on a \emph{sufficient
adjustment set} of variables (\(X\)). As a result, the ACE is expressed
as
\(\tau = E_{X}\left[E[Y_{i}| T_{i}=1, X] - E[Y_{i}| T_{i}=2, X]\right]\),
where \(E_{X}[\cdot]\) denotes the marginal expected value over \(X\)
\citep{Morgan_et_al_2014}.

@@

The structural approach, however, offers a notable advantage over the
potential outcomes approach by allowing the graphical and formal
representation of a system through directed acyclic graphs
\citep[DAG,][]{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020}.
DAGs function as heuristics, effectively conveying the presumed causal
structure of the system underlying the ACE, referred to as a
\emph{scientific model}. They do not represent detailed statistical
models but allow researchers to deduce which statistical models can
provide valid causal inferences, assuming the causal structure depicted
in the DAGs are accurate \citep{McElreath_2020}. The
Identification-Estimation flowchart in Figure~\ref{fig-IEflow} visually
represents the process of transitioning from estimands to estimates, as
well as the application of the scientific model and data to identify and
estimate causal effects.

@@

DAGs offer two key advantages for modeling causal structures. Firstly,
they represent causal relations in a nonparametric and fully interactive
manner. This feature allows for feasible causal analysis strategies
without needing the specification of the type of data or the nature of
the functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. This feature enables the
decomposition of complex structures into basic building blocks,
facilitating the analysis of these structures by focusing on the causal
assumptions associated with individual building blocks
\citep{McElreath_2020}. These building blocks can be represented in
three ways: the magnified representation, the standard representation,
and the structural causal model form \citep[SCM,][]{Morgan_et_al_2014}.

@@

The motivating example in Section~\ref{sec-framework-structural} can be
used to further illustrate how the five fundamental building blocks help
construct a system's causal structure. In this scenario, an experiment
cannot be conducted but the investigator still aims to determine
whether, \emph{on average}, there is a difference in conceptual
understanding of a topic \((Y=\{0,1\})\) between two groups of students
\((T=\{1,2\})\), described by the estimand \(E[Y| T=1] - E[Y| T=2]\).
The problem further suggests that the country to which a student belongs
(\(X\)) may influence both \(T\) and \(Y\). Such scenarios are
plausible, especially when the teaching methods depend on software or
access to technology, which may be limited in certain countries {(maybe
an example with more impact?)}. Figure~\ref{fig-example} illustrates the
plausible causal structure of this motivating example. A detailed
examination of Figures \ref{fig-mdag_example1}, \ref{fig-sdag_example1},
and SCM \ref{fig-scm_example1} reveals the presence of at least four of
the five fundamental building blocks. The figures display multiple
descendants, as indicated by pairwise relations such as
\(X \rightarrow T\), \(X \rightarrow Y\), and \(T \rightarrow Y\).
Additionally, the figures feature multiple pairs of unconnected nodes,
evident from the relations \(e_{T} \perp\!\!\!\perp e_{X}\),
\(e_{T} \perp\!\!\!\perp e_{Y}\), and \(e_{X} \perp\!\!\!\perp e_{Y}\).
Finally, the figures illustrate the fork \(X \rightarrow \{T,Y\}\), and
two colliders with \(\{X,e_{T}\} \rightarrow T\) and
\(\{X,T,e_{Y}\} \rightarrow Y\).
